{\small} High Performance P$^{3}$M N-body code: CUBEP3M

### 2012-06-08 Initial.tex
Included force tests over 10 steps

This code was tested by comparing simulations and theoretical predictions
for the effect of local primordial non-Gaussianity on the halo mass 
function and matter power spectrum (Desjacques, Seljak \& Iliev 2009). 
It has also been used to quantify the impact of local non-Gaussian initial
conditions on the halo power spectrum \citep{2009MNRAS.396...85D,
2010PhRvD..81b3006D} and bispectrum \citep{2010MNRAS.406.1014S},
 as well as the matter bispectrum \citep{2011arXiv1111.6966S}.
Fig. \ref{fig:init} shows the late time power spectrum of two RANGER4000 realizations that started off the same initial power spectrum, 
but one of which had non-Gaussian initial conditions.
We see that the difference between the two power spectra is at the sub-per cent level, but that the non-Gaussian realization
agrees much better with one loop perturbation theory \citep{XXX}.

>>>>

This code was tested by comparing simulations and theoretical predictions
for the effect of local primordial non-Gaussianity on the halo mass 
function and matter power spectrum (Desjacques, Seljak \& Iliev 2009). 
It has also been used to quantify the impact of local non-Gaussian initial
conditions on the halo power spectrum \citep{2009MNRAS.396...85D,
2010PhRvD..81b3006D} and bispectrum \citep{2010MNRAS.406.1014S},
 as well as the matter bispectrum \citep{2011arXiv1111.6966S}.
Fig. \ref{fig:init} shows the late time power spectrum of two RANGER4000 realizations that started off the same initial power spectrum, 
but one of which had non-Gaussian initial conditions.
We see that the difference between the two power spectra is at the sub-per cent level, but that the non-Gaussian realization
agrees much better with one loop perturbation theory \citep{2004PhRvD..69j3513S,2008PhRvD..78l3534T}.

### 2012-06-08 Introduction.tex
Included force tests over 10 steps

The paper is structured as follow: section \ref{sec:structure} reviews the structure and flow of the main code;
section \ref{sec:Poisson} describes how Poisson equation is solved on the two-mesh system;
we then present in section \ref{sec:scaling} the scaling of the code to very large problems,
including much larger runs that were produced at various high performance computing centers;
sections \ref{sec:accuracy} and \ref{sec:systematics} discuss the accuracy and systematic effects of the code respectively.
 We then describe in section \ref{sec:halo} the run-time halo finder, in section \ref{sec:extensions} various extensions
 to the default configuration, and conclude afterwards.

>>>>

The paper is structured as follow: section \ref{sec:structure} reviews the structure and flow of the main code;
section \ref{sec:Poisson} describes how Poisson equation is solved on the two-mesh system;
we then present in section \ref{sec:scaling} the scaling of the code to very large problems,
including much larger runs that were produced at various high performance computing centers;
section \ref{sec:accuracy}  discuss the accuracy and systematic effects of the code.
 We then describe in section \ref{sec:halo} the run-time halo finder, in section \ref{sec:extensions} various extensions
 to the default configuration, and conclude afterwards.

### 2012-06-08 Systematics.tex
Included force tests over 10 steps

\section{Accuracy}
\label{sec:accuracy}

@@

The actual force of gravity in the P$^3$M algorithm,
as felt by a single particle, is presented in Fig. \ref{fig:den_force_ppext0}.
This shows the force versus distance -- in fine cell units -- a calculation that was performed from a CITA256 realization in two steps: 
1- we compute the force on each particle in a given time step.
2- we remove a selected particle, compute the force again on all particles, and record on file the 
force difference (before  and after the removal) as a function of the distance to the `hole'.

@@

Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial and tangential directions.
It is larger than in {\small PMFAST} at grid scales, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC -- which shows a larger scatter about the theoretical value. 
As discuss in the next section, to minimize this effect, we apply a random offset to all the particles,
such that this undesired scatter is averaged out over a few time steps.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
  \caption{Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    This particular calculation was obtained in a CITA256 realization with  a box size of $500 h^{-1}\mbox{Mpc}$,
    in a single time step. In a full simulation run, the scatter averages over many time steps, thanks to the inclusion of a random offset that is imposed on each particle. 
    \label{fig:den_force_ppext0}}
\end{center}
\end{figure}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
  \caption{Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
  This scatter is significantly reduced with the random offset, as discussed in section \ref{sec:systematics}.
    \label{fig:den_force_fracErr}}
\end{center}
\end{figure}

\section{Systematics}
\label{sec:systematics}

\subsection{Mesh force at grid distances}

@@

We ensure that, on average, this solution balances out the mesh feature,
by tuning the force kernels such as to provide  force as evenly balanced as possible at grid cell distances
and at the cutoff length ($r_{c}=16$ fine cells).
These adjustments are performed from the pairwise force test described in section \ref{sec:accuracy}.
We note that this is one of the driving argument to extend the pp force outside the fine mesh cell,
since the scattering of the NGP force about the actual $1/r^{2}$ law drops rapidly as the distance increases.
As discussed in section \ref{subsec:extendedpp}, this gain in accuracy comes at a price,
a choice that  must be carefully balanced.

@@

{\bf (Show how averaging over $~sim10$ time steps reduces the scatter.)}

>>>>

\section{Accuracy and Systematics}
\label{sec:accuracy}

 This section describes the systematics effects that are inherent to our P$^{3}$M algorithm.
 We start with a demontration of the accuracy with a power spectrum measurement of a RANGER4000 simulation. 
 The halo mass function also assess the code capacity to model gravitational collapse, 
 but depends on the particular halo finder used, which we discuss in section \ref{sec:halo}.
 For the moment, we focus on the two-point function, and further quantify the accuracy of the force calculation
 with comparisons to Newton's law of gravitation.

 \subsection{Power Spectrum}
 \label{subsec:powerspectrum}

@@

\subsection{Mesh force at grid distances}
\label{subsec:force}

The actual force of gravity in the P$^3$M algorithm,
as felt by a single particle, is presented in the left panel of Fig. \ref{fig:den_force_fracErr}.
The left panel shows the force versus distance -- in fine cell units -- a calculation that was performed from a CITA256 realization in two steps: 
1- we compute the force on each particle in a given time step.
2- we remove a selected particle, compute the force again on all particles, and record on file the 
force difference (before  and after the removal) as a function of the distance to the `hole'.

@@

The right panel of Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial and tangential directions, calculated from a single time step.
It is larger than in {\small PMFAST} at grid scales, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC -- which shows a larger scatter about the theoretical value. 
This change of interpolation is essential for the pp part.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
  %\caption{Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    %This particular calculation was obtained in a CITA256 realization with  a box size of $500 h^{-1}\mbox{Mpc}$,
    %in a single time step. In a full simulation run, the scatter averages over many time steps, thanks to the inclusion of a random offset that is imposed on each particle. 
 %   \label{fig:den_force_ppext0}}
%\end{center}
%\end{figure}
%
%\begin{figure}%[ht]
 % \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
  \caption{({\it left}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    This particular calculation was obtained in a CITA256 realization with  a box size of $500 h^{-1}\mbox{Mpc}$,
    in a single time step. 
    ({\it right}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
  In a full simulation run, the scatter averages over many time steps, 
    thanks to the inclusion of a random offset that is imposed on each particle, as discussed in the main text.
    \label{fig:den_force_fracErr}}
\end{center}
\end{figure*}

@@

We ensure that, on average, this solution balances out the mesh feature,
by tuning the force kernels such as to provide  force as evenly balanced as possible at grid cell distances
and at the cutoff length ($r_{c}=16$ fine cells).
These adjustments are performed from a pairwise force test described in MPT.
We note that this is one of the driving argument to extend the pp force outside the fine mesh cell,
since the scattering of the NGP force about the actual $1/r^{2}$ law drops rapidly as the distance increases.
As discussed in section \ref{subsec:extendedpp}, this gain in accuracy comes at a price,
a choice that  must be carefully balanced.

@@

To illustrate how well the random offset performs, we present in Fig. \ref{fig:den_force_N10}
the impact of averaging over 10 time steps. We recast the same force test in which we dig a hole, 
with the addition that we apply a random offset the whole particle population at the end of the calculation.
We repeat this operation ten times, always extracting the same particle, and average the mean force and the deviation.
When compared with a single time step, we see that the scatter has reduced by almost a factor of three in the radial direction, 
while the transverse deviation shows a milder improvement.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N1.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N10.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N1.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N10.eps}
  \caption{({\it top}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$.
    ({\it bottom}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
    The left panels are calculated from a single time step, while the right panels show the average over time time steps.
    \label{fig:den_force_N10}}
\end{center}
\end{figure*}

### 2012-06-08 paper.tex
Included force tests over 10 steps

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST} that
is now one of the fastest highest performance public N-body code. 
Gravity is solved  on a two-level mesh with sub-grid precision due to the particle-particle (pp)
interactions inside and across neighboring cells.
Among important changes since {\small PMFAST},  {\small CUBEP3M} has a volumetric decomposition,
allowing for a massive parallelization over more than 20,000 cores, and has achieved close to ideal weak scaling
even at this scale.  In addition, force kernels have been improved for enhanced matching of the two mesh contributions near the cutoff length, and the code is now equipped with a unique particle identification tag system that allows reconstruction of individual particle trajectories between time steps.
In parallel, many utilities and extensions have been added to the standard code package, 
including a runtime halo finder, a tunable range for the exact pp force, a non-Gaussian initial conditions generator, etc.
We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
of the code and its utilities, when applicable.
\end{abstract}

>>>>

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST} that
is now one of the fastest highest performance public N-body code. 
Gravity is solved  on a two-level mesh with sub-grid precision due to the particle-particle (pp)
interactions inside and across neighboring cells.
Among important improvements,  {\small CUBEP3M} has a volumetric decomposition,
allowing for a massive parallelization over more than 20,000 cores, and has achieved close to ideal weak scaling
even at this scale.  In addition, force kernels have been improved for enhanced matching of the two mesh contributions near the cutoff length, and the code is now equipped with a unique particle identification tag system that allows reconstruction of individual particle trajectories between time steps.
In parallel, many utilities and extensions have been added to the standard code package, 
including a runtime halo finder, a tunable range for the exact pp force, a non-Gaussian initial conditions generator, etc.
We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
of the code and its utilities, when applicable.
\end{abstract}
